### 2016-06-01 Chapters/response_reviewers.tex

\documentclass[11pt]{article}
\usepackage{amsmath,amscd,amssymb,amstext,epsfig,textcomp}
\usepackage{lscape}
%\usepackage{multirow}
\usepackage[letterpaper]{geometry}
\geometry{verbose,tmargin=1in,bmargin=1in,lmargin=1in,rmargin=1in}
\usepackage[authoryear]{natbib}

\usepackage{color}
\newcommand{\tr}[1]{\textcolor{red}{#1}}

\begin{document}

\title{\bf Response to Reviewers of "Dynamical allocation of cellular resources as an optimal control problem: Novel
insights into microbial growth strategies" (PCOMPBIOL-S-15-02269)}

\author{Nils Giordano$^{\text{}1,2}$, Francis Mairet$^{\text{}3}$, Jean-Luc Gouz\'{e}$^{\text{}3}$, \\ Johannes Geiselmann$^{\text{}1,2,*}$, Hidde de Jong$^{\text{}2,*}$ \\ \\
1. Universit\'e Grenoble Alpes, Laboratoire Interdisciplinaire de Physique (CNRS UMR 5588) \\ 140 rue de la physique BP 87, 38402 Saint Martin d'H\`{e}res  France\\
2. INRIA, Grenoble - Rh\^one-Alpes research centre,\\ 655 avenue de l'Europe, Montbonnot, 38334 Saint Ismier Cedex, France \\
3. Inria, Sophia-Antipolis M\'{e}diterran\'{e}e research centre, 2004 route des Lucioles, BP 93,\\ 06902 Sophia-Antipolis Cedex, France \\
\\
$^{*}$ Corresponding authors with equal contributions: \\ Johannes Geiselmann (Hans.Geiselmann@ujf-grenoble.fr), \\ Hidde de Jong (Hidde.de-Jong@inria.fr)
}

\date{}

\maketitle
%\nocite{*}

\noindent  We would like to thank the reviewers for their comments on our manuscript "Dynamical allocation of cellular resources as an optimal control problem: Novel insights into microbial growth strategies" (PCOMPBIOL-S-15-02269). In this document we describe how our revised manuscript addresses the comments of reviewer 3, which mostly concern the presentation of our work. In summary, we have made the following two major changes:

\begin{enumerate}
\item The second part of the \textit{Results} section has been streamlined and we have tried to clarify the presentation throughout;
\item A \textit{Methods} section has been added to the paper and part of the material in the Supplementary Information has been shifted to this new section.
\end{enumerate}
Below we respond to the reviewer comments in detail and we indicate which changes have been made to the manuscript (major changes in red). The reviewer comments are in italic, and our response in default font.

\section*{Reviewer 3}

\textit{In this paper the authors study a resource allocation problem to identify strategies that microbes use to adapt their proteome in response to nutrient upshifts. The paper addresses a very relevant topic in systems biology with a unique (and very much needed) angle from a dynamic optimization framework. Their main results say basically that optimal regulatory strategies that maximize steady state growth are not necessarily optimal in dynamic environments. As far as I know this is the first study that reveals this property from dynamic resource allocation problems. They go further and propose a number of feedback strategies, based on precursor-sensing, nutrient-sensing, and the imbalance between precursors and ribosomes (akin to the ppGpp system), and compare them in terms of their dynamic responses. This is a very interesting paper and a solid contribution. The results seem correct but I have a number of comments to streamline ideas and homogenize the quality of the different sections of the paper in terms of language, figures and flow of ideas.} \\

\textit{- Some parts of Text S1 and S4 should be included as methods or an appendix in the main text to improve readability. I also found a small typo in S1.10, a missing \^ in the jacobian.} \\

\noindent\textbf{Answer:} We have added a \textit{Methods} section to the paper, with the following subsections:

\begin{enumerate}
\item \textit{Steady-state analysis of model.} This subsection summarizes the major results of the model analysis in Text~S1, notably explicit expressions for the optimal growth rate at steady state referred to in the main text. The details of the derivation of these expression remain in Text~S1. 
\item \textit{Model fitting.} This subsection describes how the self-replicator model was fitted against the data of \cite{Scott2010}. The contents of this subsection were formerly included in Text~S2.
\item \textit{Solution of optimal control problem.} This subsection provides an outline of the way in which the analytical results for the optimal control problem were obtained. The details of the derivation remain in Text~S3. The subsection also describes how the optimal control problem was solved numerically for generating Fig.~4, a text formerly included in Text~S3.
\item \textit{Specification and analysis of control strategies.} This subsection shows that all three control strategies discussed in the paper drive the self-replicator to the optimal steady state after the upshift. The contents were moved here from Text~S4, which has been renamed to \textit{Kinetic model of the ppGpp system in Escherichia coli} to better cover the remaining material.
\end{enumerate}

The typo in S.1.10 was corrected.\\

\textit{- Fig 2 is important but a bit convoluted. I am not sure panel 2B adds too much information and moreover, the references to Fig 2 in the text are a bit confusing, as they go back and forth between panels and it is cumbersome to follow.} \\

\noindent\textbf{Answer:} We have reduced the information overload by splitting Fig.~2 in the original manuscript into two new figures. Fig.~2 in the revised manuscript (corresponding to Fig.~2\text{A-B} in the original manuscript) presents the analysis of the self-replicator model of bacterial growth, whereas Fig.~3 in the revised manuscript (corresponding to Fig.~2\text{C-D} in the original manuscript) shows that the model is consistent with the steady-state data of Scott \textit{et al.} We believe it is important to show the dynamics in the phase plane, but we have improved the legibility of the plot (Fig.~2\textit{A} in the revised manucript). The order of the figures and the figure panels now corresponds to the order in which they are mentioned in the text.\\

\textit{- I suggest to unify the language across the whole manuscript. For example, some sections use a natural language while others a control-theoretic jargon (eg p 10), which I think it’s not the best choice unless concepts are clearly defined. A similar problem happens with several concepts in the manuscript, eg. the authors refer to the ‘environmental variable’ $e_{m}(t)$, but the constant turnover rate $E_m$ as a parameter; another example is the use of ‘intensive’ variables, vs ‘extensive’ ones, without an explanation of what these mean.} \\

\noindent\textbf{Answer:} We have tried to unify the language across the manuscript and avoid jargon when possible. In particular, we have removed some terminology specific to control theory from the main text, such as "singular arc" and "bang arc", and we have summarized the properties of the solution of the optimal control problem more informally in the main text. The notions of "bang-bang-singular" and "Pontryagin maximum principle" have been preserved, as they are central to the paper and have been used before in systems biology and mathematical biology (see the references in the \textit{Discussion}). The \textit{Methods} subsection \textit{Solution of optimal control problem} provides a more technical description and uses the appropriate terminology from control theory.

Whereas $e_m(t)$ is a variable in the general model, it becomes a constant $e_m$ (or $E_M$ in its rescaled form) in the specific upshift scenarios we consider. This has been further emphasized in the text, notably in the sections \textit{Solution of the optimal control problem} and \textit{Simple feedback control strategies}. We use the notion of "intensive" variables only once, but in a context where it is difficult to replace by another term and where it is immediately followed by an intuitive explanation and an appropriate reference (\textit{Introduction}).\\

\textit{- The introduction, first results sections and discussion are excellently written, with clean explanations and model assumptions clearly laid out. However, from p10 the flow is a bit broken and results are not so cleanly presented. I would advise the authors to streamline the text, lay out assumptions clearly (as in the first results section) and improve the figures as well. A figure for the different feedback schemes in p11 would help to clarify what is being sought, while Fig 4 could be removed or converted to an inset, as it contains little information. Another idea is to improve Figs 5 and 6: I am not sure the vector fields in panels A-B speak out the point the authors make in the text. Maybe these panels should be together for better comparison, or the authors could better explain what they mean by ‘performance’ when they compare the different dynamic responses in 5c-d vs 6c-d.} \\

\noindent\textbf{Answer:} We have rewritten the final three \textit{Results} sections to improve the overall flow of the text. In particular, we have added more intuitions and motivations, and we have tried to make the assumptions more explicit. We have also relegated Fig.~4 to a Supplementary figure (Fig.~S1) and added a new Fig.~5 in which the control strategies are graphically superposed on the self-replicator system of Fig.~1. We have preserved Figs~5 and 6 (Figs~6 and 7 in the revised manuscript), but we have better explained in the text how the vector fields in panels \textit{A-B} relate to the temporal response curves in panels \textit{C-D}.  \\

\textit{- Text S3. I think the authors could expand this section to cleanly explain the variable definitions, the derivations and pontryagin’s conditions in more detail. 3 examples: a) the constancy of the hamiltonian appears as a condition only in the middle of the derivation, b ) the definition of ‘overtaking optimality’ is missing (but crucial to the derivation, is this the role of $\lambda_0$ in the hamiltonian? ), c) the footnote 2 about abnormal extremal trajectories needs more explanation. Since this is a supplementary file, the authors can afford to explain things in detail while keeping it accessible to someone not entirely familiar with the concepts from optimal control.} \\

\noindent\textbf{Answer:} Text~S3 has been extended to provide more explanations. In particular, the conservation of the Hamiltonian along extremal trajectories is now explicitly stated earlier in the text, the notion of "overtaking optimality" is explicity defined, and footnote~2 is further developed in the text.\\

\textit{- In Fig 3B, it is unclear why the switching function has constant sign but $\alpha(t)$ switches right after $\hat{t}=2$.} \\

\noindent\textbf{Answer:} In the revised manuscript, this point is further developed at the end of the section \textit{Solution of the optimal control problem}, when discussing the figure (Fig.~4\textit{B} in the revised manuscript). We explain that the optimal solution does cross the switching curve at the moment that $\alpha$ switches from 1 to 0, around $\hat{t}=2$, and remains close to the switching curve until the next switch occurs.   \\

\textit{- In S4.1, it is unclear why ‘we know from Eq. S1.20 that $g(\hat{p})$ is the only control law...'} \\

\noindent\textbf{Answer:} The formulation was misleading: the statement is not "known", but shown here, immediately after Eq.~S4.1, with the help of Eq. S1.20. We changed the text to better bring this out (notice that this section has been moved to the \textit{Methods} section of the main text, as described above, while the proof is now included in Text~S1). \\

\textit{- S4.3, I suggest to include the definitions/notation from Bosdriesz et al explicitly to make it a standalone document.}

\noindent\textbf{Answer:} Done, in Text~S4. \\

\begin{thebibliography}{10}

\bibitem[{Scott {\it et~al\/}(2010)Scott, Gunderson, Mateescu, Zhang \&
  Hwa}]{Scott2010}
Scott M, Gunderson CW, Mateescu EM, Zhang Z, Hwa T (2010) {Interdependence of
  cell growth and gene expression: origins and consequences.} {\it Science\/}
  {\bf 330}: 1099--1102

\end{thebibliography}

\end{document}

### 2016-06-01 Chapters/core_label.tex
First meeting with H&H

For this reason, growth-rate maximization is a central hypothesis in a number of recent theoretical studies of resource allocation using coarse-grained models of the cell~\cite{molenaar_shifts_2009,scott_interdependence_2010,scott_emergence_2014}.
The models deliberately reduce the molecular complexity of regulatory networks so as to focus on generic explanatory principles~\cite{servedio_not_2014}.
Along these lines, Molenaar \textit{et al.} developed a series of simple models of the microbial cell, taking into account that growth requires the synthesis of proteins playing a role in metabolism (transporters, enzymes) and gene expression (ribosomes), in varying proportions.
Allocation parameters that maximize the growth rate were shown to account, at least in a qualitative way, for the variation of the amount of ribosomal protein as a fraction of total protein in different growth media, and for the occurrence of overflow metabolism above certain growth rates~\cite{molenaar_shifts_2009}.
Using another coarse-grained model of the cell, centered on amino acid supply (metabolism) and demand (protein synthesis), Scott \textit{et al.} derived empirical growth laws with linear relations between the ribosomal protein fraction and the growth rate, in conditions where the nutrient supply or demand are altered~\cite{scott_interdependence_2010,scott_emergence_2014}.
In their model, maximization of growth rate requires maximization of amino acid flux and is achieved for a specific, unique value of the ribosomal protein fraction.
Based on a structurally similar model, Maitra and Dill related optimal resource allocation to the basic constants of the metabolic and gene expression machinery, in particular energy efficiency~\cite{maitra_bacterial_2014}.

The assumption of growth-rate maximization may lead to correct predictions in some situations, but ignores the regulatory mechanisms achieving resource allocation and therefore cannot provide a causal explanation of cellular behavior~\cite{kremling_understanding_2015}.
Several studies have used coarse-grained models to understand which control strategies microorganisms employ to achieve (optimal) resource allocation~\cite{bosdriesz_how_2015,scott_emergence_2014,weisse_mechanistic_2015}.
Scott \textit{et al.} have shown that a robust feedforward control strategy, based on the sensing of the amino acid pool size and the corresponding adjustment of the fraction of ribosomes producing ribosomal proteins, allows the ribosomal protein fraction to be maintained close to its optimal value under a variety of growth conditions~\cite{scott_emergence_2014}.
The authors suggest that this control strategy involves the signalling molecule ppGpp, in agreement with conclusions drawn from a recent kinetic model of the regulatory mechanisms achieving optimal adjustment of the ribosomal protein fraction~\cite{bosdriesz_how_2015}.
Wei{\ss}e \textit{et al.} also developed a coarse-grained model of microbial growth based on resource allocation trade-offs~\cite{weisse_mechanistic_2015}. 
Without including specific regulatory interactions, the model accounts for the above-mentioned bacterial growth laws, predicts host-circuit interactions in synthetic biology, and relates gene regulation to the nutrient composition of the medium.

The above studies consider resource allocation at steady state, where all intensive variables describing the growing microbial culture, in particular the concentrations of its molecular components, are constant (see~\cite{fishov_microbial_1995} for a precise definition of steady-state growth and the closely related notions of balanced and exponential growth).
This requires an environment to be stable over a long period of time.
Such conditions can be achieved in the laboratory~\cite{borirak_molecular_2014}, but many microorganisms naturally experience frequently-changing conditions.
For example, \textit{E. coli} can cycle between two distinct habitats, the mammalian intestine and the earth's surface (water, sediment, soil)~\cite{savageau_escherichia_1983}.
The bacteria transit through different microenvironments in the intestinal system, where they encounter different mixes of sugars~\cite{savageau_demand_1998}.
They are even more challenged in the open environment outside the host, with a greatly fluctuating availability of carbon and energy sources and a large variability in temperature, osmolarity, oxygen, and microbial communities~\cite{blount_unexhausted_2015,vanelsas_survival_2011}.

@@

The aim of this study is to address the above fundamental questions in a specific dynamical growth scenario, namely a transition between two steady states following an environmental perturbation.
In particular, we consider the upshift of a microbial culture from a medium supporting growth at a low rate to a medium supporting growth at a high rate~\cite{ehrenberg_mediumdependent_2012}.
We develop a coarse-grained model of the cell, inspired by the self-replicator model of Molenaar \textit{et al.}~\cite{molenaar_shifts_2009}, and reformulate our questions in the context of optimal control theory~\cite{stengel_optimal_1994} to identify control schemes maximizing biomass production over an interval of time, the dynamical equivalent of growth-rate maximization.

We show that Pontryagin's Maximum Principle suggests that optimal resource allocation after a growth transition is achieved by a bang-bang-singular control law~\cite{stengel_optimal_1994}, a conjecture confirmed by direct numerical optimization.
This optimal solution provides a gold standard against which possible control strategies of the cell can be compared.
We consider simple strategies that drive the system to the steady state enabling growth at the maximal rate in the new medium, after the upshift.
In a dynamical growth scenario, the strategy sensing the concentration of precursor metabolites emerges as the best candidate, consistent with the analysis of Scott \textit{et al.} that feedforward activation of the rate of synthesis of ribosomal proteins, involving ppGpp-mediated sensing of the amino acid pool~\cite{dalebroux_ppgpp_2012,potrykus_pppgpp_2008,hauryliuk_recent_2015}, is the key regulatory mechanism for growth control.
It is possible, however, to define a strategy approaching the theoretical optimum even more closely by exploiting information on both the precursor concentration and the abundance of the gene expression machinery.
Interestingly, a thorough analysis of the functioning of the ppGpp system, as described by a kinetic model of the synthesis and degradation of this signalling molecule, suggests similarities between our two-variable control strategy and the regulation of the transcription of ribosomal RNA by ppGpp~\cite{bosdriesz_how_2015}.

@@

\begin{figure}[!h]
%\includegraphics[scale=1]{./Fig/Fig1.eps}
\caption[Self-replicator model of bacterial growth]{
{\bf Self-replicator model of bacterial growth.} External substrates $S$ enter the cell and are transformed into precursors $P$ through the action of the metabolic machinery $M$.
The precursors are used by the gene expression machinery $R$ to make the proteins composing both the metabolic machinery (transporters, enzymes, ...) and the gene expression machinery itself (RNA polymerase, ribosomes, ...).
$\alpha$ ($1-\alpha$) is the mass proportion of precursors converted into $R$ ($M$).
Thick arrows denote reactions and thin, dashed arrows denote catalytic activities.
The rate of synthesis of precursors and the rate of synthesis of proteins from precursors are denoted by $V_M$ and $V_R$, respectively.
}
\label{fig:self_replicator}
\end{figure}

@@

The dynamics of the self-replicator in Fig.~\ref{fig:self_replicator} can be described by the following system of ordinary differential equations (see \ref{S1_Text} for the derivation):
\begin{eqnarray}
\frac{dp}{dt} &=& v_M(s,r) - v_R(p,r) \, (1+\beta\, p), \label{eq:pdef}\\
\frac{dr}{dt} &=& v_R(p,r) \, (\alpha(t) - \beta\, r), \label{eq:rdef}
\end{eqnarray}
where $s$ [g~L\textsuperscript{-1}] denote the (extracellular) concentration of substrate.
$v_M(s,r)$ [g~L\textsuperscript{-1}~h\textsuperscript{-1}] and $v_R(p,r)$ [g~L\textsuperscript{-1}~h\textsuperscript{-1}] denote the precursor synthesis rate and the macromolecule synthesis rate, respectively.
The growth rate $\mu$ [h\textsuperscript{-1}] of the replicator system is defined as the relative increase of the volume, and can be rewritten with Eqs~\ref{eq:pdef}-\ref{eq:rdef} as proportional to the macromolecule synthesis rate (\ref{S1_Text}):
\begin{equation}\label{eq:growthrate}
\mu = \frac{1}{\mathtt{Vol}} \frac{d\mathtt{Vol}}{dt} = \frac{1}{M+R}\frac{d(M+R)}{dt} = \beta\, v_R(p,r).
\end{equation}

@@

The system of Eqs~\ref{eq:pdef}-\ref{eq:rdef} thus has four parameters ($e_M$, $k_R$, $K_R$, $\beta$), one of which characterizes the input from the environment ($e_M$).
The order of magnitude of the parameters can be inferred from data in the literature, as explained in \ref{S2_Text}.
Below we use the following values for the parameters $e_M = 3.6$~h\textsuperscript{-1}, $k_R = 3.6$~h\textsuperscript{-1}, $K_R=1$~g~L\textsuperscript{-1}, $\beta = 0.003$~L~g\textsuperscript{-1} (\ref{S1_Table}).
However, it should be emphasized that the conclusions of this paper do not depend on the exact quantitative values of these parameters.

@@

The nullcline for $r$ is given by $r=0$, $r=\alpha/\beta$, and $p=0$, while the nullcline for $p$ is defined by

\[
r = \frac{e_M}{\beta \left(e_M + k_R\, \frac{p}{K_R+p}\, (1+\beta p)\right)}.
\]
The nullclines define a single stable steady state $(p^*, r^*)$ (Fig.~\ref{fig:model_analysis}\textit{A} and \textit{Methods}).
At this steady state, the growth rate is constant and denoted by $\mu^*$.
The nullcline for $p$ is defined by the environment $e_M$.
The nullcline for $r$, and thus the location of the steady state with the associated growth rate, are given by $\alpha$.
Fig.~\ref{fig:model_analysis}\textit{B} shows the dependency of the steady-state growth rate $\mu^*$ on the resource allocation parameter $\alpha$.
As can be seen, $\mu^*$ is maximal for a specific, unique value of $\alpha$, which we denote $\alpha_{opt}^*$.
That is, the model predicts that there is a single optimal way to divide the precursor flux over the synthesis of the gene expression machinery and the metabolic machinery.
The same result, using a similar model, was obtained by Scott \textit{et al.}~\cite{scott_emergence_2014}.
The self-replicator model is simple enough to derive an algebraic expression for computing $\alpha^*_{opt}$ and the corresponding maximal growth rate $\mu^*_{opt}$ (\textit{Methods} and \ref{S1_Text}), which will simplify analysis of the system in later sections.

\begin{figure}[!h]
%\includegraphics[scale=1]{./Fig/Fig2.eps}
\caption[Analysis of self-replicator model of bacterial growth]{\textbf{Analysis of self-replicator model of bacterial growth.}\\
\textit{A:}~Phase-plane analysis of the self-replicator model of Eqs~\ref{eq:pdef} and \ref{eq:rdef}.
The nullclines for $p$ and $r$ are shown as solid and dashed curves, respectively.
Parameter values are $e_M = 3.6$~h\textsuperscript{-1}, $k_R = 3.6$~h\textsuperscript{-1}, $K_R=1$~g~L\textsuperscript{-1}, $\beta = 0.003$~L~g\textsuperscript{-1}, $\alpha = 0.45$.
\textit{B:}~Dependence of the growth rate at steady state $\mu^{*}$ on the resource allocation parameter $\alpha$, for two different environmental conditions (solid line, $e_M = 4.76$~h\textsuperscript{-1}; dashed line, $e_M = 1.57$~h\textsuperscript{-1}, other parameter values are $k_R = 2.23$~h\textsuperscript{-1}, $K_R=1$~g~L\textsuperscript{-1}, and $\beta = 0.003$~L~g\textsuperscript{-1}).
The maximal growth rate is attained for a unique $\alpha$, called $\alpha^*_{opt}$.
}
\label{fig:model_analysis}
\end{figure}

In order to validate the model, we verified that it can account for data on the macromolecular composition of \textit{E. coli} at steady state~\cite{scott_interdependence_2010}.
When optimizing $\alpha$ for different values of $e_M$ (assuming cells attain maximal growth), the model predicts a relation between $\alpha^*_{opt}$ and $\mu^*_{opt}$ (colored dots and black dashed line in Fig.~\ref{fig:model_validation}\textit{A}) that is quasi-linear for high growth rates.
We compared this prediction with the results of experiments where the relation between the growth rate and the mass ratio of total RNA and protein was determined in different growth media (Fig.~\ref{fig:model_validation}\textit{B}).
In the framework of our model, different media correspond to different values of $e_M$, and different total RNA/protein mass ratios to different values of $\alpha$ (up to a conversion factor), allowing a direct comparison of the model predictions in Fig.~\ref{fig:model_validation}\textit{A} with the data in Fig.~\ref{fig:model_validation}\textit{B} (see \textit{Methods}).
As can be seen, the model is able to account for the observed quasi-linear relation between the growth rate and the total mass ratio of RNA and protein. 
Moreover, for realistic values of $k_R$ and $e_M$, a good quantitative fit is obtained (\textit{Methods} and \ref{S1_Table}).

>>>>

For this reason, growth-rate maximization is a central hypothesis in a number of recent theoretical studies of resource allocation using coarse-grained models of the cell~\cite{molenaar_shifts_2009,scott_interdependence_2010,scott_emergence_2014}.
The models deliberately reduce the molecular complexity of regulatory networks so as to focus on generic explanatory principles~\cite{servedio_not_2014}.
Along these lines, Molenaar \etal developed a series of simple models of the microbial cell, taking into account that growth requires the synthesis of proteins playing a role in metabolism (transporters, enzymes) and gene expression (ribosomes), in varying proportions.
Allocation parameters that maximize the growth rate were shown to account, at least in a qualitative way, for the variation of the amount of ribosomal protein as a fraction of total protein in different growth media, and for the occurrence of overflow metabolism above certain growth rates~\cite{molenaar_shifts_2009}.
Using another coarse-grained model of the cell, centered on amino acid supply (metabolism) and demand (protein synthesis), Scott \etal derived empirical growth laws with linear relations between the ribosomal protein fraction and the growth rate, in conditions where the nutrient supply or demand are altered~\cite{scott_interdependence_2010,scott_emergence_2014}.
In their model, maximization of growth rate requires maximization of amino acid flux and is achieved for a specific, unique value of the ribosomal protein fraction.
Based on a structurally similar model, Maitra and Dill related optimal resource allocation to the basic constants of the metabolic and gene expression machinery, in particular energy efficiency~\cite{maitra_bacterial_2014}.

The assumption of growth-rate maximization may lead to correct predictions in some situations, but ignores the regulatory mechanisms achieving resource allocation and therefore cannot provide a causal explanation of cellular behavior~\cite{kremling_understanding_2015}.
Several studies have used coarse-grained models to understand which control strategies microorganisms employ to achieve (optimal) resource allocation~\cite{bosdriesz_how_2015,scott_emergence_2014,weisse_mechanistic_2015}.
Scott \etal have shown that a robust feedforward control strategy, based on the sensing of the amino acid pool size and the corresponding adjustment of the fraction of ribosomes producing ribosomal proteins, allows the ribosomal protein fraction to be maintained close to its optimal value under a variety of growth conditions~\cite{scott_emergence_2014}.
The authors suggest that this control strategy involves the signalling molecule ppGpp, in agreement with conclusions drawn from a recent kinetic model of the regulatory mechanisms achieving optimal adjustment of the ribosomal protein fraction~\cite{bosdriesz_how_2015}.
Wei{\ss}e \etal also developed a coarse-grained model of microbial growth based on resource allocation trade-offs~\cite{weisse_mechanistic_2015}. 
Without including specific regulatory interactions, the model accounts for the above-mentioned bacterial growth laws, predicts host-circuit interactions in synthetic biology, and relates gene regulation to the nutrient composition of the medium.

The above studies consider resource allocation at steady state, where all intensive variables describing the growing microbial culture, \tr{in particular the concentrations of its molecular components, are constant (see~\cite{fishov_microbial_1995} for a precise definition of steady-state growth and the closely related notions of balanced and exponential growth).}
This requires an environment to be stable over a long period of time.
Such conditions can be achieved in the laboratory~\cite{borirak_molecular_2014}, but many microorganisms naturally experience frequently-changing conditions.
For example, \textit{E. coli} can cycle between two distinct habitats, the mammalian intestine and the earth's surface (water, sediment, soil)~\cite{savageau_escherichia_1983}.
The bacteria transit through different microenvironments in the intestinal system, where they encounter different mixes of sugars~\cite{savageau_demand_1998}.
They are even more challenged in the open environment outside the host, with a greatly fluctuating availability of carbon and energy sources and a large variability in temperature, osmolarity, oxygen, and microbial communities~\cite{blount_unexhausted_2015,vanelsas_survival_2011}.

@@

The aim of this study is to address the above fundamental questions in a specific dynamical growth scenario, namely a transition between two steady states following an environmental perturbation.
In particular, we consider the upshift of a microbial culture from a medium supporting growth at a low rate to a medium supporting growth at a high rate~\cite{ehrenberg_mediumdependent_2012}.
We develop a coarse-grained model of the cell, inspired by the self-replicator model of Molenaar \etal~\cite{molenaar_shifts_2009}, and reformulate our questions in the context of optimal control theory~\cite{stengel_optimal_1994} to identify control schemes maximizing biomass production over an interval of time, the dynamical equivalent of growth-rate maximization.

We show that Pontryagin's Maximum Principle suggests that optimal resource allocation after a growth transition is achieved by a bang-bang-singular control law~\cite{stengel_optimal_1994}, a conjecture confirmed by direct numerical optimization.
This optimal solution provides a gold standard against which possible control strategies of the cell can be compared.
We consider simple strategies that drive the system to the steady state enabling growth at the maximal rate in the new medium, after the upshift.
In a dynamical growth scenario, the strategy sensing the concentration of precursor metabolites emerges as the best candidate, consistent with the analysis of Scott \etal that feedforward activation of the rate of synthesis of ribosomal proteins, involving ppGpp-mediated sensing of the amino acid pool~\cite{dalebroux_ppgpp_2012,potrykus_pppgpp_2008,hauryliuk_recent_2015}, is the key regulatory mechanism for growth control.
It is possible, however, to define a strategy approaching the theoretical optimum even more closely by exploiting information on both the precursor concentration and the abundance of the gene expression machinery.
Interestingly, a thorough analysis of the functioning of the ppGpp system, as described by a kinetic model of the synthesis and degradation of this signalling molecule, suggests similarities between our two-variable control strategy and the regulation of the transcription of ribosomal RNA by ppGpp~\cite{bosdriesz_how_2015}.

@@

\begin{figure}[h]
\includegraphics[scale=1]{./Fig/Fig1.eps}
\caption[Self-replicator model of bacterial growth]{
{\bf Self-replicator model of bacterial growth.} External substrates $S$ enter the cell and are transformed into precursors $P$ through the action of the metabolic machinery $M$.
The precursors are used by the gene expression machinery $R$ to make the proteins composing both the metabolic machinery (transporters, enzymes, ...) and the gene expression machinery itself (RNA polymerase, ribosomes, ...).
$\alpha$ ($1-\alpha$) is the mass proportion of precursors converted into $R$ ($M$).
Thick arrows denote reactions and thin, dashed arrows denote catalytic activities.
The rate of synthesis of precursors and the rate of synthesis of proteins from precursors are denoted by $V_M$ and $V_R$, respectively.
}
\label{fig:self_replicator}
\end{figure}

@@

The dynamics of the self-replicator in Fig.~\ref{fig:self_replicator} can be described by the following system of ordinary differential equations (see Text~\ref{S1_Text} for the derivation):
\begin{eqnarray}
\frac{dp}{dt} &=& v_M(s,r) - v_R(p,r) \, (1+\beta\, p), \label{eq:pdef}\\
\frac{dr}{dt} &=& v_R(p,r) \, (\alpha(t) - \beta\, r), \label{eq:rdef}
\end{eqnarray}
where $s$ [g~L\textsuperscript{-1}] denote the (extracellular) concentration of substrate.
$v_M(s,r)$ [g~L\textsuperscript{-1}~h\textsuperscript{-1}] and $v_R(p,r)$ [g~L\textsuperscript{-1}~h\textsuperscript{-1}] denote the precursor synthesis rate and the macromolecule synthesis rate, respectively.
The growth rate $\mu$ [h\textsuperscript{-1}] of the replicator system is defined as the relative increase of the volume, and can be rewritten with Eqs~\ref{eq:pdef}-\ref{eq:rdef} as proportional to the macromolecule synthesis rate (Text~\ref{S1_Text}):
\begin{equation}\label{eq:growthrate}
\mu = \frac{1}{\mathtt{Vol}} \frac{d\mathtt{Vol}}{dt} = \frac{1}{M+R}\frac{d(M+R)}{dt} = \beta\, v_R(p,r).
\end{equation}

@@

The system of Eqs~\ref{eq:pdef}-\ref{eq:rdef} thus has four parameters ($e_M$, $k_R$, $K_R$, $\beta$), one of which characterizes the input from the environment ($e_M$).
The order of magnitude of the parameters can be inferred from data in the literature, as explained in Text~\ref{S2_Text}.
Below we use the following values for the parameters $e_M = 3.6$~h\textsuperscript{-1}, $k_R = 3.6$~h\textsuperscript{-1}, $K_R=1$~g~L\textsuperscript{-1}, $\beta = 0.003$~L~g\textsuperscript{-1} (Table \ref{S1_Table}).
However, it should be emphasized that the conclusions of this paper do not depend on the exact quantitative values of these parameters.

@@

%The dynamics of the self-replicator model can be easily analyzed in the phase plane $(p,r)$, assuming constant resource allocation $\alpha(t)=\alpha$ and a constant environment $e_M$.
The nullcline for $r$ is given by $r=0$, $r=\alpha/\beta$, and $p=0$, while the nullcline for $p$ is defined by

\[
r = \frac{e_M}{\beta \left(e_M + k_R\, \frac{p}{K_R+p}\, (1+\beta p)\right)}.
\]
The nullclines define a single stable steady state $(p^*, r^*)$ \tr{(Fig.~\ref{fig:model_analysis}\textit{A} and \textit{Methods}).}
At this steady state, the growth rate is constant and denoted by $\mu^*$.
The nullcline for $p$ is defined by the environment $e_M$.
The nullcline for $r$, and thus the location of the steady state with the associated growth rate, are given by $\alpha$.
\tr{Fig.~\ref{fig:model_analysis}\textit{B}} shows the dependency of the steady-state growth rate $\mu^*$ on the resource allocation parameter $\alpha$.
As can be seen, $\mu^*$ is maximal for a specific, unique value of $\alpha$, which we denote $\alpha_{opt}^*$.
That is, the model predicts that there is a single optimal way to divide the precursor flux over the synthesis of the gene expression machinery and the metabolic machinery.
The same result, using a similar model, was obtained by Scott \etal~\cite{scott_emergence_2014}.
The self-replicator model is simple enough to derive an algebraic expression for computing $\alpha^*_{opt}$ and the corresponding maximal growth rate $\mu^*_{opt}$ \tr{(\textit{Methods} and Text~\ref{S1_Text}),} which will simplify analysis of the system in later sections.

\begin{figure}[!ht]
\includegraphics[scale=1]{./Fig/Fig2.eps}
\caption[Analysis of self-replicator model of bacterial growth]{\textbf{\tr{Analysis of self-replicator model of bacterial growth.}}\\
\tr{\textit{A:}~Phase-plane analysis of the self-replicator model of Eqs~\ref{eq:pdef} and \ref{eq:rdef}.
The nullclines for $p$ and $r$ are shown as solid and dashed curves, respectively.
}
\tr{Parameter values are $e_M = 3.6$~h\textsuperscript{-1}, $k_R = 3.6$~h\textsuperscript{-1}, $K_R=1$~g~L\textsuperscript{-1}, $\beta = 0.003$~L~g\textsuperscript{-1}, $\alpha = 0.45$.}
\tr{
\textit{B:}~Dependence of the growth rate at steady state $\mu^{*}$ on the resource allocation parameter $\alpha$, for two different environmental conditions (solid line, $e_M = 4.76$~h\textsuperscript{-1}; dashed line, $e_M = 1.57$~h\textsuperscript{-1}, other parameter values are $k_R = 2.23$~h\textsuperscript{-1}, $K_R=1$~g~L\textsuperscript{-1}, and $\beta = 0.003$~L~g\textsuperscript{-1}).
The maximal growth rate is attained for a unique $\alpha$, called $\alpha^*_{opt}$.}
}
\label{fig:model_analysis}
\end{figure}

\begin{figure}[!ht]
\includegraphics[scale=1]{./Fig/Fig3.eps}
\caption[Self-replicator model accounts for bacterial growth laws]{
\textbf{\tr{Self-replicator model accounts for bacterial growth laws.}}\\
\tr{\textit{A:}~Predicted quasi-linear relation between the maximal growth rate $\mu^{*}_{opt}$ and the corresponding optimal resource allocation $\alpha^*_{opt}$, for different values of $e_M$ (different colors).
The colored dots indicate $\alpha^*_{opt}$ and $\mu^{*}_{opt}$  for $k_R = 2.23$~h\textsuperscript{-1} and different $e_M$, and the dashed black line the relation for all intermediate values of $e_M$.
The dashed colored lines indicate the relation between $\alpha^*_{opt}$ and $\mu^{*}_{opt}$ obtained when, for a given value of $e_M$, the value of $k_R$ is decreased (lower $k_R$ leads to lower $\mu^{*}_{opt}$).
The solid grey curves correspond to $(\mu^*,\alpha)$-profiles like those shown in Fig.~\ref{fig:model_analysis}\textit{B}.
\textit{B:}~Measured relation between the total RNA/protein mass ratio and the growth rate, in different growth media with different doses of a translation inhibitor (data from~\cite{scott_interdependence_2010}).
For each medium, indicated by a color, five different concentrations of inhibitor were used (higher dose leads to lower growth rate).
Growth-medium compositions are given in the original publication and error bars represent standard deviations.
The dashed black and colored lines are the same as in panel \textit{A}, indicating the good quantitative correspondence between model predictions and experimental data for the chosen parameter values, obtained by fitting the model to the data points (see \textit{Methods} for details).
}
}
\label{fig:model_validation}
\end{figure}
